Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-02-05 analysis/paper/paper.Rmd
add sentences and a figure for spatial analysis in dicussion section and method section

```{r metricCVstest}
library(cvequality)
# read metric data
kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
                                        "KWL-pottery-metrics-tidy-ave.csv"))
# number for each phase 
kwl_p_ave_total <-
  kwl_p_ave %>% 
  count(period)
# inline code
metric_all <- sum(kwl_p_ave_total$n)
pre <- kwl_p_ave_total$n[3]
euro <- kwl_p_ave_total$n[2]
chi <- kwl_p_ave_total$n[1]

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis. Most of them cover every vessel component from rim to bottom. To fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. Also, we explore spatial distribution of pottery to examine if clustering changes over time across the sampling area. As craft specialization increases, pottery distribution might present a shift from dispersed clusters to concentrated clusters that hints the emergence of specialized group working in workshop.

<!-- we also look at spatial distribution to see if clustering changes over time... ...as craft specialization increases, we might expect spatial clustering to increase... as possible indicator of craft specialization -->

@@

saveRDS(jpgs_imported_coo,
        here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
```

@@

# Mean pottery shapes of three phases
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))

@@

# Isodeformation lines
png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)

@@

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables.

>>>>

```{r metricCVstest}
library(cvequality)
# read metric data
kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
# number for each phase 
kwl_p_ave_total <-
  kwl_p_ave %>% 
  count(period)
# inline code
metric_all <- sum(kwl_p_ave_total$n)
pre <- kwl_p_ave_total$n[3]
euro <- kwl_p_ave_total$n[2]
chi <- kwl_p_ave_total$n[1]

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis. Most of them cover every vessel component from rim to bottom. The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. In addition, we use Monte Carlo method to test for randomness in spatial locations of pots to explore whether the distribution is meaningful indicated by clustered or dispersed pattern. As craft specialization increases, pottery distribution might present a shift from dispersed pattern to concentrated clusters that hints the emergence of specialized groups.

@@

saveRDS(jpgs_imported_coo,
        here("analysis/data/derived_data/jpgs_imported_coo.rds"))
```

@@

# Mean pottery shapes of three phases
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here("analysis/data/derived_data/jpgs_imported_coo.rds"))

@@

# Isodeformation lines
png(here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)

@@

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables. In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence. However, the distribution of pottery is more similar to the random distributions during the Chinese period. This contradicts to our expectation that clustered patten will be observed with an increase in pottery standardization since the emergence of specialized groups. Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.

```{r Monte-Carlo-spatial-pattern, fig.cap= "Histograms of simulated average nearest-neighbour distances (ANN) values from 1000 simulations for three phases. X-axis values based on meteres represent ANN expected value. Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}
knitr::include_graphics(here("analysis", "figures", "plot-kde-ann-histograms.jpg"))
```
